Confronting preferential sampling in count and occupancy surveys: diagnosis and model-based triage \footnotemark[2]

### 2015-11-20 manuscript/pref_sampling.tex
Changes to simple spatial model in TMB (got rid of intercept), a few ms changes
Draft simulations conducted and written up in ms

\author{Paul B. Conn\affil{1}\corrauth, Devin S. Johnson\affil{1}, James T. Thorson\affil{2}}

@@

One of the main advantages of using SDMs advanced in the literature is that one is no longer beholden to predetermined sampling frames, and can potentially use data gathered from non-randomized designs or platforms of opportunity to make inferences about animal populations \citep{JohnsonEtAl2010}.  However,
in a recent paper, \citet{DiggleEtAl2010} emphasized that spatially explicit statistical models can easily provide biased estimates when sampling is nonrandom. The potential for biased estimates arises when sampling locations disproportionately target locations where the response of interest is higher (or lower) than the mean response than would be predicted from explanatory covariates with complete knowledge of the system.  In the context of SDMs, this might occur if sampling disproportionately occurs in locations where animals are known to be present or of high abundance, regardless of predictive covariates. For example, if volunteer inventory participants have access to multiple sites with similar covariate values, bias might arise if they consistently choose sites where species are thought or known to be present.  Bias might also arise if surveying effort is higher near bases of operations, and if animal abundance is higher (or lower) near bases of operations than elsewhere in the landscape.

@@

By contrast, CSPS (hereafter, PS), the primary focus of this article, arises when the locations being sampled and the process of interest (e.g. density, occupancy) are conditionally dependent given modeled covariates \citep{DiggleEtAl2010}.  For instance, PS can occur when the investigator uses a priori knowledge or observations of the state variable obtained during sampling to allocate survey effort in places where abundance or occurrence is known to be high. \citet{DiggleEtAl2010} showed that this type of preferential sampling can lead to bias when this extra information is not included in models for the state variable of interest.  Specifically, PS arises when we consider the set of sampled locations as stochastic and when $[{\bf R},{\bf Z} | {\bf x}] \neq [{\bf R}|{\bf x}][{\bf Z}|{\bf x}]$ \citep{DiggleEtAl2010}.  We use this definition of PS throughout the rest of the manuscript, noting that it is somewhat different than has sometimes been used in the SDM literature.  For instance, \citet{MerckxEtAl2011} use the term ``preferential sampling" to refer to the process of visiting some sites more often than others, while \citet{ManceurKuhn2014} define it as occurring when the locations selected for sampling are a function of an environmental covariate.  Neither of these latter conditions are problematic outside of the specialized field of presence-only modelling.

@@

As in the pollution example, having good explanatory covariates may also reduce bias when fitting SDMs to animal encounter data under PS.  However, in many ecological applications, predictive covariates are only able to explain a portion of variation present in the data.  If the locations selected for sampling are related (intentionally or unintentionally) to some unmodelled factor related to abundance, bias may still occur.  Despite the clear potential for bias in SDMs, we have been unable to find many cases where PS \citep[\textit{sensu}][]{DiggleEtAl2010} is discussed with regard to SDMs.  For instance,
\citet{ChakrabortyEtAl2010} acknowledged the likely presence of PS when fitting SDMs to data obtained using nonrandomized designs, but did not attempt to model it.

@@

Given these limitations, our present task is to generalize PS models to the types of data more typical of SDMs, and to allow the degree of PS to vary across the landscape.  Like \citet{DiggleEtAl2010} and \citet{PatiEtAl2011}, we impose a joint model for the process of interest (animal abundance or occurence) and the locations chosen for sampling. For the process model, we start with Eq. \ref{eq:process} as a general formulation for non-Gaussian data.  We then expand the link-scale expectation from this model
(i.e. Eq. \ref{eq:mu}) as follows:
\begin{eqnarray}
  \mu_i & = & \beta_0 + {\bf x}_i \boldsymbol{\beta}+\delta_i+\epsilon_i,
\label{eq:mu_aug}
\end{eqnarray}
where $\delta_i$ is a spatially referenced random effect.  Next, we model the binary indicator for sample inclusion using a Bernoulli distribution:
\begin{eqnarray}
  R_i & \sim & \text{Bernoulli}(h^{-1}(\nu_i)),
\end{eqnarray}
where $h()$ denotes a link function appropriate for binary data (e.g. logit, probit).  We then write the linear predictor for this model as
\begin{eqnarray}
  \nu_i & = & \beta_0^* + {\bf x}_i^* \boldsymbol{\beta}^*+\eta_i+{\bf B} \delta_i + \varepsilon_i.
\label{eq:nu}
\end{eqnarray}
 In a similar fashion to the model for the state process, the sampling intensity model has an intercept ($\beta_0^*$), explanatory covariates (${\bf x}_i^*$), fixed effect regression parameters ($\boldsymbol{\beta}^*$), spatially autocorrelated random effects ($\eta_i$ and $\delta_i$), and normally distributed error $\varepsilon_i$.  The predictive covariates ${\bf x}_i$ from Eq. \ref{eq:mu_aug} and ${\bf x}_i^*$ from Eq. \ref{eq:nu} need not be the same (although they can be).  Note also that the  spatially autocorrelated random effect $\delta_i$ is included in both Eqs. \ref{eq:mu_aug} and \ref{eq:nu}, allowing for dependency in the two models, with the matrix {\bf B} describing the strength and type of dependence between the sampling process and underlying density.

 The formulation in Eq. \ref{eq:nu} is the same as previously proposed by other authors for hierarchical multivariate models with spatial dependence \citep[cf.][]{RoyleBerliner1999}.  There are multiple ways of structuring ${\bf B}$ depending on the complexity with which one wants spatial dependence that is desired \citep{RoyleBerliner1999}.  For instance, setting ${\bf B} = {\bf O}_{S \times S}$ corresponds to an absence of spatial dependence (and thus no preferential sampling).  Setting ${\bf B} = b {\bf I}$, where $b$ is an estimated parameter and {\bf I} is an $(S \times S)$ identity matrix corresponds to the linear preferential sampling model suggested by \citet{DiggleEtAl2010} and \citet{PatiEtAl2011}.  Alternatively, we could allow the degree of PS to vary across the landscape.  For instance, one can contemplate a trend surface model for preferential sampling by specifying a diagonal matrix for ${\bf B}$, with entries given by $b_0 + b_1 \textrm{lat}_i + b_2 \textrm{long}_i$, where $b_0$, $b_1$, and $b_2$ are estimated parameters and $\textrm{lat}_i$ and $\textrm{long}_i$ give latitude and longitude, respectively \citep{RoyleBerliner1999}.  Theoretically, one could include more highly parameterized structures for spatial dependence, such as higher order trend surfaces or a two dimensional spline \citep{RoyleBerliner1999,UnpublishedManuscript}, but the ability to robustly estimate the parameters of such a model is likely dependent on having a rich, spatially balanced dataset, which is often not the case in ecological applications.

@@

To illustrate PS, we conducted a count survey simulation experiment.  For each of 100 simulations, we generated abundance of a hypothetical species over a $30 \times 30$ grid as a function of two spatially autocorrelated covariates (with log-linear effects of covariates and an interaction term; Appendix S1).   For each simulated landscape we generated count data using Eqns \ref{eq:process} \& \ref{eq:obs_pois}, setting $A_i = p_i = 1.0$ for demonstration purposes.  We conducted three types of surveys of $K=50$ survey units: (i) a spatially balanced survey, (ii) an unequal probability design where the probabilities of sample inclusion was a function of covariate values (see Appendix S1), and an unequal probability design where the probability of sample inclusion was set to $(Z_i+1)/\sum(Z_i+1)$ (representing PS).  We provided a count-based process model with the first of the covariates as an explanatory variable, treating the second as a variable conspiring to influence distribution and abundance and imparting spatial autocorrelation but which remains unknown to the investigator. We used Markov chain Monte Carlo to conduct statistical inference, generating posterior predictions as $N = \sum_i Z_i$ and compared the performance (bias, precision, 90\% credible interval coverage) of these predictions relative to known, true values.

\subsection{MYSTERY SPECIES X}

@@

@@

@@

\begin{figure*} %[ht]
\begin{center}
\includegraphics[width=170mm]{PS_maps.pdf}
\caption{An example of a single simulation replicate examining the effect of preferential sampling on estimates of abundance from a species distribution model.  First, two spatially autocorrelated covariates are generated (panels A-B).  Second, true abundance is generated conditional on these covariates (C).  Next, three hypothetical sampling designs are employed to generate animal counts, including a spatially balanced sample (D), a design where sampling is proportional to values of a modeled covariate (E), and where sampling is preferential - i.e. more likely to occur where abundance is high (F).  Finally, spatially explicit estimates of abundance are generated using a traditional SDM to each of the count datasets (G-I).  Although the map produced in (I) does not look particularly bad in this instance, preferential sampling led to estimates of abundance that were biased 20\% high.}
\label{fig:PS_sims}
\end{center}
\end{figure*}

>>>>

\author{Paul B. Conn\affil{1}\corrauth, James T. Thorson\affil{2}, Devin S. Johnson\affil{1}}

@@

One of the main advantages of using SDMs advanced in the literature is that one is no longer beholden to predetermined sampling frames, and can potentially use data gathered from non-randomized designs or platforms of opportunity to make inferences about animal populations \citep{JohnsonEtAl2010}.  However,
in a recent paper, \citet{DiggleEtAl2010} emphasized that spatially explicit statistical models can easily provide biased estimates when sampling is nonrandom. The potential for biased estimates arises when sampling locations disproportionately target locations where the response of interest is higher (or lower) than expected given a particular set of explanatory covariates.  In the context of SDMs, this might occur if sampling disproportionately occurs in locations where animals are known to be present or of high abundance, regardless of predictive covariates. For example, if volunteer inventory participants have access to multiple sites with similar covariate values, bias might arise if they consistently choose sites where species are thought or known to be present.  Bias might also arise if surveying effort is higher near bases of operations, and if animal abundance is higher (or lower) near bases of operations than elsewhere in the landscape.

Need to acknowledge that "sample selection bias" has been addressed extensively in SDM modeling with presence-only data (e.g. adjusting "background" or "availability" data to be similar to the locations chosen for sampling).

@@

By contrast, CSPS (hereafter, PS), the primary focus of this article, arises when the locations being sampled and the process of interest (e.g. density, occupancy) are conditionally dependent given modeled covariates \citep{DiggleEtAl2010}.  For instance, PS can occur when the investigator uses a priori knowledge or observations of the state variable obtained during sampling to allocate survey effort in places where abundance or occurrence is known to be high. \citet{DiggleEtAl2010} showed that this type of preferential sampling can lead to bias when this extra information is not included in models for the state variable of interest.  Specifically, PS arises when we consider the set of sampled locations as stochastic and when $[{\bf R},{\bf Z} | {\bf x}] \neq [{\bf R}|{\bf x}][{\bf Z}|{\bf x}]$ \citep{DiggleEtAl2010}, where \textbf{R} is an indicator vector taking on a value of $R_i = 1.0$ if sampling unit $i$ is sampled and is zero otherwise.  We use this definition of PS throughout the rest of the manuscript, noting that it is somewhat different than has sometimes been used in the SDM literature.  For instance, \citet{MerckxEtAl2011} use the term ``preferential sampling" to refer to the process of visiting some sites more often than others, while \citet{ManceurKuhn2014} define it as occurring when the locations selected for sampling are a function of an environmental covariate.  Neither of these latter conditions are problematic outside of the specialized field of presence-only modelling.

@@

As in the pollution example, having good explanatory covariates may also reduce bias when fitting SDMs to animal encounter data under PS.  However, in many ecological applications, predictive covariates are only able to explain a portion of variation present in the data.  If the locations selected for sampling are related (intentionally or unintentionally) to some unmodelled factor related to abundance, bias may still occur.  Despite the clear potential for bias in SDMs, we have been unable to find many cases where PS \citep[\textit{sensu}][]{DiggleEtAl2010} is discussed with regard to SDMs.  One such example is
\citet{ChakrabortyEtAl2010}, who acknowledged the likely presence of PS when fitting SDMs to data obtained using nonrandomized designs, but did not attempt to model it.

@@

Given these limitations, our present task is to generalize PS models to the types of data more typical of SDMs, and to allow the degree of PS to vary across the landscape.  Like \citet{DiggleEtAl2010} and \citet{PatiEtAl2011}, we impose a joint model for the process of interest (animal abundance or occurence) and the locations chosen for sampling. For the process model, we start with Eq. \ref{eq:process} as a general formulation for non-Gaussian data.  We then expand the link-scale expectation from this model
(i.e. Eq. \ref{eq:mu}) as follows:
\begin{eqnarray}
  \mu_i & = & \beta_0 + {\bf x}_i \boldsymbol{\beta}+\delta_i,
\label{eq:mu_aug}
\end{eqnarray}
where $\delta_i$ is a spatially referenced random effect.  Next, we model the binary indicator for sample inclusion using a Bernoulli distribution:
\begin{eqnarray}
 \label{eq:R}
  R_i & \sim & \text{Bernoulli}(h^{-1}(\nu_i)),
\end{eqnarray}
where $h()$ denotes a link function appropriate for binary data (e.g. logit, probit).  We then write the linear predictor for this model as
\begin{eqnarray}
  \label{eq:nu}
  \nu_i & = & \beta_0^* + {\bf x}_i^* \boldsymbol{\beta}^*+\eta_i+{\bf B} \delta_i.
\label{eq:nu}
\end{eqnarray}
 In a similar fashion to the model for the state process, the sampling intensity model has an intercept ($\beta_0^*$), explanatory covariates (${\bf x}_i^*$), fixed effect regression parameters ($\boldsymbol{\beta}^*$), spatially autocorrelated random effects ($\eta_i$ and $\delta_i$), and normally distributed error $\varepsilon_i$.  The predictive covariates ${\bf x}_i$ from Eq. \ref{eq:mu_aug} and ${\bf x}_i^*$ from Eq. \ref{eq:nu} need not be the same (although they can be).  Note also that the  spatially autocorrelated random effect $\delta_i$ is included in both Eqs. \ref{eq:mu_aug} and \ref{eq:nu}, allowing for dependency in the two models, with the matrix {\bf B} describing the strength and type of dependence between the sampling process and underlying density.

 The formulation in Eq. \ref{eq:nu} is the same as previously proposed by other authors for hierarchical multivariate models with spatial dependence \citep[cf.][]{RoyleBerliner1999}.  There are multiple ways of structuring ${\bf B}$ depending on the complexity of spatial dependence desired for the preferential sampling process \citep{RoyleBerliner1999}.  For instance, setting ${\bf B} = {\bf O}_{S \times S}$ corresponds to an absence of spatial dependence (and thus no preferential sampling).  Setting ${\bf B} = b {\bf I}$, where $b$ is an estimated parameter and {\bf I} is an $(S \times S)$ identity matrix corresponds to the linear preferential sampling model suggested by \citet{DiggleEtAl2010} and \citet{PatiEtAl2011}.  Alternatively, we could allow the degree of PS to vary across the landscape.  For instance, one can contemplate a trend surface model for preferential sampling by specifying a diagonal matrix for ${\bf B}$, with entries given by $b_0 + b_1 \textrm{lat}_i + b_2 \textrm{long}_i$, where $b_0$, $b_1$, and $b_2$ are estimated parameters and $\textrm{lat}_i$ and $\textrm{long}_i$ give latitude and longitude, respectively \citep{RoyleBerliner1999}.  Theoretically, one could include more highly parameterized structures for spatial dependence, such as higher order trend surface or spline formulation \citep{RoyleBerliner1999}, but the ability to robustly estimate the parameters of such a model is likely dependent on having a rich, spatially balanced dataset, which is often not the case in ecological applications.

@@

To illustrate PS and demonstrate that our proposed model has reasonable performance, we conducted a small simulation experiment.  For each of 100 simulations, we generated abundance of a hypothetical species over a $25 \times 25$ grid as
\begin{eqnarray*}
  N_i \sim \textrm{Poisson}(\mu_i),
\end{eqnarray*}
where $i$ indexes survey unit $i$, and $\mu_i$ is determined according to Eq. \ref{eq:mu_aug}. Abundance was generated as a function of a single spatially autocorrelated landscape covariate, as well as residual spatial autocorrelation ($\delta_i$) and overdispersion (fig. \ref{PS_sims}). Specific details of of data generation procedures are presented in Appendix S1.

For each simulated landscape we generated three virtual count surveys using eqs. \ref{eq:R} and \ref{eq:nu}.  Each survey had $\boldsymbol{\beta}^*=\eta_i=0$ (that is, not covariate of spatially autocorrelated random effects), but differed in how the matrix ${\bf B}$ was parameterized.  In the first, we set ${\bf B}=0$, so that the survey was a simple random sample.  For the second and third, we set ${\bf B}$ to be a diagonal matrix with entries $b=1$ and $b=5$, respectively, so that the probability of sampling a given survey unit (grid cell) was explicitly dependent on the latent abundance in that unit.  We refer to these scenarios as moderate and pathological preferential sampling, respectively (see fig. \ref{fig:sim_R}).  Simulations were configured so that $n=50$ of the 625 survey units were sampled; each survey was set to cover half of the target cell.

We fitted two different models to each count dataset, both of which were provided the habitat covariate used (in part) to generate the data for which a log-linear coefficient $\beta$ was estimated.  In the first estimation model, the elements of \textbf{B} in eq. \ref{eq:nu} were all set to zero.  In this case, the abundance and sampling process submodels were independent, as is the case canonical SDMs (at lest when fitted to presence-absence or count data).  In the second estimation model, we included an explicit connection between the distribution of animal abundance and the sampling process by setting $\textbf{B}=b\textbf{I}$, where $b$ is an estimated parameter, and \textbf{I} is an identity matrix.

We used maximum a posteriori (MAP) estimation to conduct statistical inference. This approach is Bayesian; prior parameter distributions are specified and inference is conducted relative to marginal posterior distributions. However, in contrast to Markov chain Monte Carlo (MCMC) where the posterior is sampled via simulation, the joint posterior is maximized numerically in this case.  We used Template Model Builder \citep[TMB; ][]{KristensenEtAl2015}, interfaced with the R programming environment, to conduct maximization. The TMB software uses a Laplace approximation to integrate out random effects, and a bias correction algorithm \citep{Tierney1989} to obtain abundance estimates that properly account for uncertainty attributable to random effects (\textit{correct, Jim?}).  This approach allowed for a facile implementation and fast computing times, allowing us to conduct simulation and model testing with greater efficiency than would have been possible with MCMC.  In this study, we report the results of 500 simulation replicates.  Further detail on statistical methods are provided in Appendix S1; requisite R and TMB code will be published to a publicly accessible repository upon acceptance, and is also available at \url{https://github.com/NMML/pref_sampling/}.

\subsection{Bearded seal aerial surveys}

\subsection{Avian point counts}

@@

Estimates of cumulative animal abundance across simulated landscapes were median unbiased for both estimation methods when the sites selected for sampling were independent of animal density, though when $b$ was estimated abundance estimates were more right skewed and had higher variance (fig. \ref{fig:sim_bias}).  Under moderate preferential sampling ($b=1$), the jointly modeling the density and the sampling process (i.e., estimating $b$) led to a median bias of 5\%, while the canonical SDM model had a median bias of 40\%.  Under pathological preferential sampling ($b=5$), both estimation methods were extremely biased, but was even more severe for the naive model ignoring preferential sampling (fig. \ref{sim_bias}).

@@

Presumably, trends less biased when pref sampling ignored.

@@

\begin{figure*} %[ht]
\begin{center}
\includegraphics[width=170mm]{Pref_samp_sim_maps.pdf}
\caption{An example of a single simulation replicate examining estimates of abundance from a naive species distribution model under preferential sampling.  First, true abundance (C) is generated as a function of a spatially autocorrelated covariate (A) and a spatially autocorrelated random effect (B). Second, counts are generated for three different types of surveys, including a simple random sample ($b=0$; D) and surveys with moderate ($b=1$; E) or pathological ($b=5$; F) levels of preferential sampling.  Finally, spatially explicit estimates of abundance are generated using a traditional SDM (with $b$ set to 0.0) to each of the count datasets (G-I).  In this particular simulation replicate, cumulative abundance was underestimated by 18\% when $b=0$, overestimated by 17\% when $b=1$, and overestimated by 293\% when $b=5$.  For a summary of bias over 500 simulation replicates, see fig. \ref{fig:sim_bias}.}
\label{fig:PS_sims}
\end{center}
\end{figure*}

\begin{figure*} %[ht]
\begin{center}
\includegraphics[width=170mm]{sim_R.pdf}
\caption{Expected relationship between the probability of a survey unit being selected for sampling and its abundance residual in the simulation study. The base case $b=0$ represents simple random sampling, while $b=1$ and $b=5$ represent moderate and pathological levels of preferential sampling, respectively.  Also shown are is the realized distribution (smoothed histogram) of abundance residuals among survey units in the simulation study, scaled to fit in the plot margins (solid black line).}
\label{fig:sim_R}
\end{center}
\end{figure*}

\begin{figure*} %[ht]
\begin{center}
\includegraphics[width=170mm]{bias.pdf}
\caption{Relative proportional error in abundance from the simulation experiment as computed with respect the posterior mode with a bias correction. Each boxplot summarizes the distribution of relative proportional error as a function of the type of sampling, including simple random sampling ($b=0$), moderate preferential sampling ($b=1$), and pathological preferential sampling ($b=5$). Results vary by the type of estimation model; in the ``independent" model, $b$ is set to 0.0; in the ``joint" model, $b$ is estimated.  Lower and upper limits of each box correspond to first and third quartiles, while whiskers extend to the lowest and highest observed bias within 1.5 interquartile range units from the box.  Points denote outliers outside of this range.  Horizontal lines within boxes denote median bias.  }
\label{fig:sim_bias}
\end{center}
\end{figure*}
